Linear and Differential Cryptanalysis of DES

### 2009-01-29 talk/draft1.tex
1st commit of Linear slides

\begin{frame}
Introducing SPN mathematically

@@

\subsection{Linear Attack on SPN}
\begin{frame}
\large{Basic math concepts behind the Linear attack:}\\
\vspace{5 mm}
Linear cryptanalysis tries to take advantage of high probability occurrences of  linear expressions involving plaintext bits, ciphertext bits and subkey bits.  It is a known plaintext attack.\\
\vspace{5 mm}
Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
\end{frame}

\begin{frame}
Intro to linear / affine approximation ...

\begin{frame}
Linear characteristic / approximation of SBox ...

>>>>

\begin{frame}
SPN S-Box and Permutation:\\
\vspace{5mm}
SPN uses a single 4-bit S-Box that has the following structure:
\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
\hline
Input: & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & A & B & C & D & E & F \\
\hline
Output:& E & 4 & D & 1 & 2 & F & B & 8 & 3 & A & 6 & C & 5 & 9 & 0 & 7 \\
\hline
\end{tabular}\\
\vspace{5mm}
And the following 16-bit permutation:
\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
\hline
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16\\
\hline
1 & 5 & 9 & 13 & 2 & 6 & 10 & 14 & 3 & 7 & 11 & 15 & 4 & 8 & 12 & 16 \\
\hline
\end{tabular}
\vspace{5mm}

@@

\subsection{Linear Attack on SPN}
\begin{frame}
\large{Basic Definitions for the Linear attack:}\\
\vspace{5 mm}
Linear cryptanalysis tries to take advantage of high probability occurrences of  linear expressions involving plaintext bits, ciphertext bits and subkey bits.  It is a known plaintext attack.\\
\vspace{5 mm}
Define $P_i$ where $i = 1, 2, \dots 16$ as the $i$-th plaintext bit.\\
Define $C_i$ where $i = 1, 2, \dots 16$ as the $i$-th cyphertext bit.\\
Define $U_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th input bit to the $j$-th round of SPN.\\
Define $V_{j,i}$ where $j = 1, 2, 3, 4$ and $i = 1, 2, \dots 16$ be the $i$-th output bit of the $j$-th round of SPN.\\
\end{frame}

\begin{frame}
Linear and Affine approximation of S-Box:\\
\vspace{5mm}
How do we come up with the desired expression for the entire cipher? We start by looking at the only non-linear component, the S-Box.\\
\vspace{5mm}
To find the linear/afine approximation of the S-Box we simply consider every possible expression of the input bits $X_i$ and output bits $Y_j$. Thus the expression has the form
\[ \bigoplus_{i \in U} X_i = \bigoplus_{j \in V} Y_j \]
Where $U$ and $V$ range over all possible subsets of $\{1, 2, 3 ,4\}$. We then compare how often this expression coincides with the S-Box. \\
\vspace{5mm}
Note that there are 16 possibilities for $U$ and $V$, hence 256 total possible expressions. (this is for a 4-bit S-box)
\end{frame}

\begin{frame}
S-Box approximatin example:\\
\vspace{5mm}
We can take the following expression as an example:
\[ X_2 \oplus X_3 = Y_1 \oplus Y_3 \oplus Y_4 \]
Applying all possible values for the input $X$ bits it turns out that the expression holds in 12 out of the 16 cases. Hence, this expression has a bias of $12/16 - 1/2 = 1/4$.\\
\vspace{5mm}
Similarly the following expression
\[ X_3 \oplus X_4 = Y_1 \oplus Y_4 \]
Has a bias of $-3/8$, which is higher in magnitude than the previous one. It is the magnitude of the bias that is important.
\end{frame}

\begin{frame}
The number of agreements (minus 8) between the S-Box and every possible expression is summarized in the table below. Thus to get the bias, one must only divide by 16.\\
\vspace{5mm}
\tiny
\begin{tabular}{|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|r|}
\hline
 &0&1&2&3&4&5&6&7&8&9&A&B&C&D&E&F\\
\hline
0&+8&0&0&0&0&0&0&0&0&0&0&0&0&0&0&0 \\
\hline
1&0&0&-2&-2&0&0&-2&\textbf{+6}&+2&+2&0&0&+2&+2&0&0 \\
\hline
2&0&0&-2&-2&0&0&-2&-2&0&0&+2&+2&0&0&\textbf{-6}&+2 \\
\hline
3&0&0&0&0&0&0&0&0&+2&\textbf{-6}&-2&-2&+2&+2&-2&-2 \\
\hline
4&0&+2&0&-2&-2&-4&-2&0&0&-2&0&+2&+2&-4&+2&0 \\
\hline
5&0&-2&-2&0&-2&0&+4&+2&-2&0&-4&+2&0&-2&-2&0  \\
\hline
6&0&+2&-2&+4&+2&0&0&+2&0&-2&+2&+4&-2&0&0&-2 \\
\hline
7&0&-2&0&+2&+2&-4&+2&0&-2&0&+2&0&+4&+2&0&+2  \\
\hline
8&0&0&0&0&0&0&0&0&-2&+2&+2&-2&+2&-2&-2&\textbf{-6} \\
\hline
9&0&0&-2&-2&0&0&-2&-2&-4&0&-2&+2&0&+4&+2&-2 \\
\hline
A&0&+4&-2&+2&-4&0&+2&-2&+2&+2&0&0&+2&+2&0&0 \\
\hline
B&0&+4&0&-4&+4&0&+4&0&0&0&0&0&0&0&0&0 \\
\hline
C&0&-2&+4&-2&-2&0&+2&0&+2&0&+2&+4&0&+2&0&-2 \\ 
\hline
D&0&+2&+2&0&-2&+4&0&+2&-4&-2&+2&0&+2&0&0&+2 \\
\hline
E&0&+2&+2&0&-2&-4&0&+2&-2&0&0&-2&-4&+2&-2&0 \\
\hline
F&0&-2&-4&-2&-2&0&+2&0&0&-2&+4&-2&-2&0&+2&0 \\
\hline
\end{tabular}
\small
\end{frame}

\begin{frame}
What this means for 1 Round:\\
\vspace{5mm}
Note, from the previous table, that the expression $X_3 \oplus X_4 = Y_1 \oplus Y_4$ has a bias of $-3/8$
